---
title: "practical_exercise_3, Methods 3, 2021, autumn semester"
author: '[FILL IN YOUR NAME]'
date: "[FILL IN THE DATE]"
output: html_document
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readbulk, MuMIn)
```

# Exercises and objectives
The objectives of the exercises of this assignment are:  
1) Download and organise the data and model and plot staircase responses based on fits of logistic functions  
2) Fit multilevel models for response times  
3) Fit multilevel models for count data  

REMEMBER: In your report, make sure to include code that can reproduce the answers requested in the exercises below (__MAKE A KNITTED VERSION__)  
REMEMBER: This assignment will be part of your final portfolio

## Exercise 1

Go to https://osf.io/ecxsj/files/ and download the files associated with Experiment 2 (there should be 29).  
The data is associated with Experiment 2 of the article at the following DOI https://doi.org/10.1016/j.concog.2019.03.007  

1) Put the data from all subjects into a single data frame

```{r}
df <- read_bulk(
  directory = 'experiment_2/',
          fun = read_csv
  )
```


2) Describe the data and construct extra variables from the existing variables

    i. add a variable to the data frame and call it _correct_ (have it be a _logical_ variable). Assign a 1 to each row where the subject indicated the correct answer and a 0 to each row where the subject indicated the incorrect answer (__Hint:__ the variable _obj.resp_ indicates whether the subject answered "even", _e_ or "odd", _o_, and the variable _target_type_ indicates what was actually presented.
    
```{r}
df$correct <- 2
for (i in 1:18131) {
if (df$obj.resp[i] == "o" & df$target.type[i] == "odd" | df$obj.resp[i] == "e" & df$target.type[i] == "even") {
  df$correct[i] <-  1
} else {
  df$correct[i] <-  0
}
}

df$correct <- as.logical(df$correct)
```
    
    ii. describe what the following variables in the data frame contain, _trial.type_, _pas_, _trial_, _target.contrast_, _cue_, _task_, _target_type_, _rt.subj_, _rt.obj_, _obj.resp_, _subject_ and _correct_. (That means you can ignore the rest of the variables in your description). For each of them, indicate and argue for what `class` they should be classified into, e.g. _factor_, _numeric_ etc.
    
```{r}
glimpse(df)

df$trial.type <- as.factor(df$trial.type)
  #The staircase trial types were only introduced in experiment 2 and is an adaptive procedure that allows to collect more data at the     threshold visibility. It is not explained what a staircase procedure actually is, but it is used at the beginning of the study before    collecting the actual trials.The other trials are experiment trials.

df$pas <- as.factor(df$pas)
  #The Perceptual Awareness Scale, ranging from 1-4.
  #No Experience (NE) -> 1
  #Weak Glimpse (WG) -> 2
  #Almost Clear Experience (ACE) -> 3
  #Clear Experience (CE) -> 4

# trial
  #The text said that each participant performed 864 experiment trials and 18 practice trials beforehand. In the data, it seems that 431   trials are recorded for each participant and a varying number of staircase trials

# target.contrast
  #the opacity of the target relative to the background, it was adjusted to match the threshold of each participant

df$cue <- as.factor(df$cue)
  #a cue of the possible number of digits presented for each trial, i.e. a kind of framing. Repeated 12 times before a new cue was used

df$task <- as.factor(df$task)
  #he 3 types of tasks: 
  #Singles
  #Pairs
  #Quadruplets 

df$target.type <- as.factor(df$target.type)
  #Whether the target was odd or even

# rt.subj 
  #reaction time of the subjective response

df$rt.obj <- as.numeric(df$rt.obj)
  #reaction time of the objective response

df$obj.resp <- as.numeric(df$obj.resp)
  #the subjects answer to whether the number was odd or even

df$subject <- as.factor(df$subject)
  #participantID

# correct
  #our variable indicating with a 1 that the participant answered correctly and a 0 if they answered incorrectly
```
    
    iii. for the staircasing part __only__, create a plot for each subject where you plot the estimated function (on the _target.contrast_ range from 0-1) based on the fitted values of a model (use `glm`) that models _correct_ as dependent on _target.contrast_. These plots will be our _no-pooling_ model. Comment on the fits - do we have enough data to plot the logistic functions?
    
```{r}
df_staircase <- filter(df, trial.type == "staircase")
model.1.2.3 <- glm(correct ~ target.contrast*subject, data= df_staircase, family = binomial(link = "logit"))
model.1.2.3

df_staircase$fitted_correct <- fitted(model.1.2.3)

fitted_ggplot <- ggplot(data = df_staircase, aes(x = target.contrast, y = fitted_correct)) +
  geom_point(aes(target.contrast, fitted_correct), color = "skyblue1")+
  facet_wrap(~ subject) +
  theme_bw()
fitted_ggplot
```
    
    iv. on top of those plots, add the estimated functions (on the _target.contrast_ range from 0-1) for each subject based on partial pooling model (use `glmer` from the package `lme4`) where unique intercepts and slopes for _target.contrast_ are modelled for each _subject_
    
```{r}
model.1.2.4 <- glmer(correct ~ target.contrast + (target.contrast|subject), data = df_staircase, family = binomial(link = "logit"))
model.1.2.4
summary(model.1.2.4)

df_staircase$fitted_partial <- fitted(model.1.2.4)

fitted_ggplot2 <- ggplot(data = df_staircase, aes(x = target.contrast, y = fitted_correct, color = fitted_partial)) +
  geom_point(aes(target.contrast, fitted_correct), color = "salmon1")+
  geom_point(aes(target.contrast, fitted_partial), color = "skyblue1") +
  facet_wrap(~ subject) +
  labs(x="Contrast", y="Correctness", "fitted_partial") +
  theme_bw()
fitted_ggplot2
```
    
    v. in your own words, describe how the partial pooling model allows for a better fit for each subject

  - partial pooling enables us access information across all of the participants
  - it improves our estimates throughout individuals
  - it illustrates individual variance more

## Exercise 2

Now we __only__ look at the _experiment_ trials (_trial.type_)  

1) Pick four subjects and plot their Quantile-Quantile (Q-Q) plots for the residuals of their objective response times (_rt.obj_) based on a model where only intercept is modelled  
    
```{r}
df_experiment <- filter(df, trial.type=="experiment")
df_4participants <- filter(df_experiment, subject=="001" | subject=="002" | subject=="003" | subject=="004")

model.2.1.1 <- lmer(rt.obj ~ 1 + (1 | subject) , data = df_4participants)
model.2.1.1

qqnorm(residuals(model.2.1.1))
qqline(residuals(model.2.1.1))
```
    
    i. comment on these
    
      - it's not normal, it's skewed, it has heavy tails
    
    ii. does a log-transformation of the response time data improve the Q-Q-plots?
    
```{r}
df_4participants$rt.obj.log <- log(df_4participants$rt.obj)
model.2.1.2 <- lmer(rt.obj.log ~ 1 + (1 | subject), data = df_4participants)
model.2.1.2

qqnorm(residuals(model.2.1.2))
qqline(residuals(model.2.1.2))

#it doesn't improve it, but it looks different
```
    
2) Now do a partial pooling model modelling objective response times as dependent on _task_? (set `REML=FALSE` in your `lmer`-specification)
    
```{r}
model.2.2.1 <- lmer(rt.obj ~ task + (1 | subject), df_experiment, REML=FALSE)

model.2.2.1
```
    
    i. which would you include among your random effects and why? (support your choices with relevant measures, taking into account variance explained and number of parameters going into the modelling)
    
    - we chose subject due to individual differences
    - variance explained (R squared):
    
```{r}
r.squaredGLMM(model.2.2.1)
```
    
    ii. explain in your own words what your chosen models says about response times between the different tasks
    
    - singles has the fastest response time and quadruplet is the fastest
    
3) Now add _pas_ and its interaction with _task_ to the fixed effects

```{r}
model.2.3 <- lmer(rt.obj ~ task * pas + (1 | subject), df_experiment)
```

    i. how many types of group intercepts (random effects) can you add without ending up with convergence issues or singular fits?
    
```{r}
model.2.3.1 <- lmer(rt.obj ~ task * pas + (1 | subject) + (1 | correct), df_experiment)
model.2.3.1
##we have singular fit already from 2 intercepts
```
    
    ii. create a model by adding random intercepts (without modelling slopes) that results in a singular fit - then use `print(VarCorr(<your.model>), comp='Variance')` to inspect the variance vector - explain why the fit is singular (Hint: read the first paragraph under details in the help for `isSingular`)
    
```{r}
model.2.3.2 <- lmer(rt.obj ~ task * pas + (1 | subject) + (1 | correct), df_experiment)
print(VarCorr(model.2.3.2), comp='Variance')

?isSingular

# our "dimensions" of the variance-covariance matrix have been estimated as zero
```
    
    iii. in your own words - how could you explain why your model would result in a singular fit?
    
      - because our second intercept is zero
    
## Exercise 3

1) Initialise a new data frame, `data.count`. _count_ should indicate the number of times they categorized their experience as _pas_ 1-4 for each _task_. I.e. the data frame would have for subject 1: for task:singles, pas1 was used # times, pas2 was used # times, pas3 was used # times and pas4 was used # times. You would then do the same for task:pairs and task:quadruplet  

```{r}
## you can start from this if you want to, but you can also make your own from scratch
library(dplyr) #for renaming

counting <- df %>% 
  count(subject, task, pas)

counting

##create a df out of our table
data.count <- as.data.frame.matrix(counting)

data.count <- data.count %>%
   rename(count = n)

data.count$count <-  as.numeric(data.count$count)
data.count$pas <-  as.numeric(data.count$pas)
data.count$task <-  as.numeric(data.count$task)
data.count$subject <-  as.numeric(data.count$subject)
```        

2) Now fit a multilevel model that models a unique "slope" for _pas_ for each _subject_ with the interaction between _pas_ and _task_ and their main effects being modelled  
    i. which family should be used?
    
```{r}
model.3.2.0 <- glmer(count ~ pas * task + (1 + pas | subject), family = poisson, data = data.count)
model.3.2.0
```
    
    ii. why is a slope for _pas_ not really being modelled?  
    iii. if you get a convergence error, try another algorithm (the default is the _Nelder_Mead_) - try (_bobyqa_) for which the `dfoptim` package is needed. In `glmer`, you can add the following for the `control` argument: `glmerControl(optimizer="bobyqa")` (if you are interested, also have a look at the function `allFit`)
    iv. when you have a converging fit - fit a model with only the main effects of _pas_ and _task_. Compare this with the model that also includes the interaction  
    v. indicate which of the two models, you would choose and why  
    vi. based on your chosen model - write a short report on what this says about the distribution of ratings as dependent on _pas_ and _task_  
    vii. include a plot that shows the estimated amount of ratings for four subjects of your choosing 
3) Finally, fit a multilevel model that models _correct_ as dependent on _task_ with a unique intercept for each _subject_  
    i. does _task_ explain performance?
    
    ii. add _pas_ as a main effect on top of _task_ - what are the consequences of that?  
    iii. now fit a multilevel model that models _correct_ as dependent on _pas_ with a unique intercept for each _subject_
    iv. finally, fit a model that models the interaction between _task_ and _pas_  and their main effects  
    v. describe in your words which model is the best in explaining the variance in accuracy  

